A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-07-16 Manuscript/PNAS-Adams-01.tex
pnas-01

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Evolutionary biologists wish to quantify and compare the strength of
phylogenetic signal across traits, but analytical tools for these
comparisons are generally lacking. Here we develop a standardized effect
size based on \(\kappa\), (\(Z_\kappa\)), which measures the strength of
phylogenetic signal on a common statistical scale, and provides a
mechanism for formally comparing the strength of phylogenetic signal
across datasets. Additionally, we find that a commonly used parameter
(Pagel's \(\lambda\)) is unsuitable for this purpose. Our procedure
enables biologists to quantitatively address hypotheses that compare the
strength of phylogenetic signal between various phenotypic traits, even
when those traits are found in different evolutionary lineages or have
different units or scales.}

@@

\begin{abstract}
Macroevolutionary studies frequently characterize the phylogenetic
signal in phenotypes, and wish to compare the strength of that signal
across traits. However, analytical tools for such comparisons have
largely remained underdeveloped. Here we evaluate the efficacy of one
commonly used parameter (Pagel's \(\lambda\)) to estimate the strength
of phylogenetic signal in phenotypic traits, and evaluate the degree to
which \(\lambda\) correctly identifies known levels of phylogenetic
signal. We find that \(\lambda\) behaves as a Bernoulli random variable,
and that estimates are increasingly skewed at larger and smaller input
levels of phylogenetic signal. Further, the precision of \(\lambda\) in
estimating actual levels of phylogenetic signal is often inaccurate, and
biological interpretations of the strength of phylogenetic signal based
on \(\lambda\) are therefore compromised. As an alternative, we propose
a standardized effect size based on \(\kappa\), (\(Z_\kappa\)), which
measures the strength of phylogenetic signal more reliably than does
\(\lambda\), and places that signal on a common scale for statistical
comparison. We develop tests based on \(Z_\kappa\) to provide a
mechanism for formally comparing the strength of phylogenetic signal
across datasets, in much the same manner as effect sizes may be used to
summarize patterns in quantitative meta-analysis. Our approach extends
the phylogenetic comparative toolkit to address hypotheses that compare
the strength of phylogenetic signal between various phenotypic traits,
even when those traits are found in different evolutionary lineages or
have different units or scales.
\end{abstract}

@@

Investigating macroevolutionary patterns of trait variation requires a
phylogenetic perspective, because the shared ancestry among species
violates the assumption of independence among trait values that is
common for statistical tests (1, 2). Accounting for this evolutionary
non-independence is the purview of \emph{phylogenetic comparative
methods} (PCMs): a suite of analytical tools that condition trends in
the data on the phylogenetic relatedness of observations (3--12). These
methods are predicated on the notion that phylogenetic signal -- the
tendancy for closely related species to display similar trait values --
is present in cross-species datasets (1, 13, 14). Indeed, under numerous
evolutionary models, phylogenetic signal is to be expected, as
stochastic character change along the hierarchical structure of the tree
of life generates trait covaration among related taxa (1, 14, 15).

Several analytical tools have been developed to quantify phylogenetic
signal in phenotypic datasets (13, 14, 16--19), and their statistical
properties -- namely type I error rates and statistical power -- have
been investigated to determine under what conditions phylogenetic signal
can be detected (15, 18, 20--25). One of the most widely used methods
for characterizing phylogenetic signal is Pagel's \(\lambda\) (13),
which transforms the lengths of the internal branches of the phylogeny
to improve the fit of data to the phylogeny via maximum likelihood (13,
26). When incorporated in PGLS, \(\lambda\) serves as a tuning parameter
which is optimized via log-likelihood profiling while evaluating the
covariation between the dependent and independent variables, given the
phylogeny (13, 26). To infer whether phylogenetic signal differs from no
signal or a Brownian motion model of evolutionary divergence, the
observed model fit using \(\hat\lambda\) may be statistically compared
to that using \(\lambda=0\) or \(\lambda=1\) via likelihood ratio tests
(26--28) or confidence limits (29). Another widely used measure is
Blomberg's \(\kappa\) (14), which measures the ratio of observed trait
variation to the amount expected under Brownian motion. Blomberg's
\(\kappa\) can be treated as a test statistic by employing a permutation
test to generate its sampling distribution (14, 18) for determining
whether significant phylogenetic signal is present in data. Both
\(\lambda\) and \(\kappa\) seem intuitive to interpret, as a value of
\(0\) for both corresponds to no phylogenetic signal, while a value of
\(1\) corresponds to the amount of phylogenetic signal expected under
Brownian motion. Thus, it is tempting to regard both \(\lambda\) and
\(\kappa\) as descriptive statistics that measure the relative strength
of phylogenetic signal, providing an estimate of its magnitude for
comparison.

The appeal of Pagel's \(\lambda\) and Blomberg's \(\kappa\) as
descriptive statistics is that they provide a basis for interpreting
``weak'' versus ``strong'' phylogenetic signal; i.e., small versus large
values of \(\hat{\lambda}\) or \(\kappa\), respectively, in a
comparative sense (30--32). Nonetheless, an important question that has
yet to be considered is whether these statistics are or can be converted
to effect sizes for comparative analyses across datasets? To be
statistics representing phylogenetic signal, they should have reliable
distributional properties, which could be revealed with simulation
experiments. For instance, as a proportional random variable bounded by
\(0\) and \(1\), we might expect that \(\hat{\lambda}\) follows a
Bernoulli distribution (\textbf{add ref}); i.e., branch lengths in a
tree are scaled proportionally to the probability that data arise from a
BM process. Given a known \(\lambda\) value used to generate random data
on a tree, we would also expect that the mean of an empirical sampling
distribution of \(\hat{\lambda}\) would approximately equal \(\lambda\);
the dispersion of \(\hat{\lambda}\) would be largest at intermediate
values of \(\lambda\), \(\hat{\lambda}\) would be predictable over the
range of \(\lambda\) with respect to treesize; the distribution of
\(\hat{\lambda}\) would be symmetric at intermediate values of
\(\lambda\) and more skewed toward values of 0 or 1; and that the
distribution of \(\hat{\lambda}\) will be more platykurtic at
intermediate values of \(\lambda\), becoming more leptokurtic toward 0
and 1 (\textbf{add same ref}). Prior work (20) seems to support some of
these conjectures, based superficially on statistical moments for a
given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of
ref. (20)). However, because the ``strength of Brownian motion'' was
simulated as a varied weighted-average of data simulated on trees with
\(\lambda=0\) and \(\lambda=1\) and not as prescribed values of
\(\lambda\) (20), interpretation of these patterns is challenging.

By contrast, Blomberg's \(\kappa\), which is positively ubounded, might
be expected to follow a normal distribution (\textbf{add ref}). Thus we
might expect that \(\kappa\) is symmetrically distributed across
different strengths of phylogenetic signal, for any \(\lambda\) used to
generate data. This attribute seemed less reasonable based on the
simulations performed by Münkemüller et al.~(20), which suggested that
distributions were positively skewed and that Blomberg's \(\kappa\)
might not behave as a statistic that follows a normal distribution.
However, because their simulations used a weighted combination of
simulated phylogenetic signal strengths, strong inferences are not
possible (and distributional attributes were not the intended result of
their simulations). Thus, for both Pagel's \(\lambda\) or Blomberg's
\(\kappa\), evaluation of statistical moments across a range of
\(\lambda\) used to generate data would be valuable for adjudicating the
relibalilty of these statistics. Furthermore, these are statistics that
appear to have expected values that vary with tree size (20), making
comparisons across studies challenging. Therefore, transformation of
these statistics into \(Z\)-scores in the same simulation experiments
would allow evaluation of the efficacy of each statistic to yield effect
sizes that could be used for comparisons of the strength of phylogenetic
signal across traits and clades.

Here we used simulation experiments to compare the distributional
attributes of \(\hat{\lambda}\) and \(\kappa\), plus their effect sizes
(\(Z\)-scores), across a range of tree size and phylogenetic signal
strength. We find that estimates of \(\hat{\lambda}\) are increasingly
skewed at larger and smaller input levels of phylogenetic signal and at
smaller tree sizes, vary widely for a given input value of \(\lambda\),
and that the precision of \(\hat{\lambda}\) is not constant across its
range. By contrast, estimates of \(\kappa\) are more consistent across
tree sizes, and are normally distributed across the range of input
levels of \(\lambda\), making \(\kappa\) a more reliable statistic. We
then propose an effect size based on \(\kappa\), (\(Z_{\kappa}\)), which
provides consistent estimates of the strength of phylogenetic signal
across tree sizes and signal strength, and facilitates quantitative
comparisons of the relative strength of phylogenetic signal across
datasets.

@@

\textbf{Lambda (\(\lambda\)) estimates of phylogenetic signal are
innacurate.} Computer simulations revealed that for \(\hat{\lambda}\),
the distributional expectations of a Bernoulli variable were mostly
upheld. First, the mean value of \(\hat{\lambda}\) did increase as
\(\lambda\) increased, but it was negatively-biased (particularly for
small tree sizes), and was consistently less than the input \(\lambda\)
value across most of its range (Fig. 1 black line). Second, the standard
deviation of \(\hat{\lambda}\) was largest at intermediate values of
\(\lambda\) and smallest at extreme values, implying that the precision
in estimating \(\lambda\) varied across the range of input values (Fig.
1 red line). Additionally, standard deviations of \(\hat{\lambda}\) were
negatively associated with tree size, and for tress of 128 species or
less, \(\hat{\lambda}\) was quite variable, except for cases when
\(\lambda\) was near or equal to \(1\). Third, the distributions of
\(\hat{\lambda}\) were not normal across its range, but became
increasingly skewed at more extreme values of \(\lambda\) (Fig. 1 blue
line). For small tree sizes, it was also clear that distributions were
more platykurtic at intermeidate values of \(\hat{\lambda}\). Taken
together these results reveal that \(\hat\lambda\) inconsistently
estimated phylogenetic signal, both across tree sizes and across the
range of input values. Additional simulations (Supplemental Information)
revealed that incorporating \(\lambda\) in PGLS anova and regression did
not adversely affect the statistical properties of model fitting (type I
error, power, parameter estimation); thus, it is reasonable to include
\(\lambda\) in PGLS as a parameter for tuning the degree of phylogenetic
signal in the dependent variables during the analysis. However, the
statistical properties revealed in Fig. 1 demonstrate that \(\lambda\)
is unsuitable as an effect size for measuring the strength of
phylogenetic signal in data.

\begin{figure}
\centering
\includegraphics{new.fig.1.temp.png}
\caption{Patterns in \(\lambda\).{}}
\end{figure}

\textbf{Kappa (\(\kappa\)) estimates of phylogenetic signal are more
stable.} Simulation results for \(\kappa\) demonstrated that this
measure displayed better statistical properties. First, as expected,
mean values of \(\kappa\) increased consistently with increasing signal
(\(\lambda\)) irrespective of tree size (Fig. 2 black line).
Additionally, the standard deviation of \(\kappa\) was consistent across
tree sizes (Fig. 2 red line), and while it increased with \(\lambda\),
it was always less than the corresponding mean. This finding is perhaps
unsurprising, as \(\kappa\) is bound by 0, and was never large for small
values of \(\lambda\). Importantly, \(\kappa\) was normally distributed
across the range of input \(\lambda\), and remained consistent in this
pattern regardless of tree size (Fig. 2 blue line). In fact,
distributions of \(\kappa\) displayed slight skewing only at small tree
sizes and when \(\lambda\) was near or equal to \(1\). This result
differs from those of (20), where skewing appears to be from combining
random values generated independently. These findings reveal that
\(\kappa\) is more reliable as an estimate of phylogenetic signal.

\begin{figure}
\centering
\includegraphics{new.fig.2.temp.png}
\caption{Patterns in \(\kappa\).{}}
\end{figure}

\textbf{Effect sizes from \(\kappa\) (\(Z_{\kappa}\)) better
characterize phylogenetic signal.} To measure the strength of
phylogenetic signal on a common scale, we derived effect sizes
(Z-scores) for both \(\lambda\) and \(\kappa\) (see Methods) and
evaluated them relative to known phylogenetic signal. Both
\(Z_{\lambda}\) and \(Z_{\kappa}\) were associated with input
phylogenetic signal (\(\lambda\)), indicating that both statistics
captured the observed signal (Fig. 3). However, effect sizes from
\(\hat{\lambda}\) made little sense, as they were more strongly
associated with tree size than they were with the actual phylogenetic
signal in the data (Fig. 3). By contrast, \(Z_{\kappa}\) was much more
consistent accross tree sizes, and exhibited a much stronger association
with phylogenetic signal strength as compared to tree size (Fig. 3).
Thus between the two statistics, \(Z_{\kappa}\) is a more reliable
measure of the strength of phylogenetic signal, and may be used to
compare levels of phylogenetic signal across datasets.

\begin{figure}
\centering
\includegraphics{new.fig.3.temp.png}
\caption{Patterns in Z.{}}
\end{figure}

\textbf{A test statistic (\(\hat{Z}_{12}\)) allows meaningful
comparisons across datasets.} To statistically compare the strength of
phylogenetic signal across datasets we derived a two-sample test
statistic (\(\hat{Z}_{12}\): see Methods). As an example of its utility,
we compared \(Z_{\kappa}\) for two ecologically-relevant traits in
plethodontid salamander (Fig. 4): surface area to volume ratios (SA:V)
and relative body width (\(\frac{BW}{SVL}\)) (33, 34). While both traits
contained significant phylogenetic signal, tests based on
\(\hat{Z}_{12}\) revealed that the degree of phylogenetic signal was
significantly stronger in SA:V (Fig. 4). Biologically, this observation
may be interpreted by the fact that tropical species -- which form a
monophyletic group within plethodontids -- display greater variation in
SA:V, which covaries with disparity in their climatic niches (34). Thus,
greater phylogenetic signal in SA:V is to be expected.

@@

\textbf{To be edited.} address hypotheses that compare the strength of
phylogenetic signal between various phenotypic traits, even when those
traits are found in different evolutionary lineages or have different
units or scales.

It is common in comparative evolutionary studies to characterize the
phylogenetic signal in phenotypic traits to determine the extent to
which shared evolutionary history has generated trait covariation among
taxa. However, while numerous analytical approaches may be used to
quantify phylogenetic signal (13, 14, 16--18), methods that explicitly
measure the strength of phylogenetic signal, or facilitate comparisons
among datasets, have remained underdeveloped. In this study, we
evaluated the precision of one common measure, Pagel's \(\lambda\), and
explored its efficacy for characterizing the strength of phylogenetic
signal in phenotypic data. Using computer simulations, we found that the
precision of \(\lambda\) increased with increasing sample sizes; a
pattern noted previously (25), and one that conformed with parametric
statistical theory (35). However, we also found that vastly different
\(\lambda\) estimates could be obtained from data containing the same
level of phylogenetic signal, and that similar \(\lambda\) estimates may
be obtained from data containing differing levels of phylogenetic
signal. Further, the precision of \(\lambda\) varied with the strength
of phylogenetic signal, where lower precision was observed when in data
whose phylogenetic signal was of intemediate strength. From these
findings we conclude that \(\lambda\) is not a reliable indicator of the
observed strength of phylogenetic signal in phenoytpic datasets, and
that biological interpretations of the strength of signal based on this
parameter may innacurately characterize such effects.

>>>>

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Evolutionary biologists wish to quantify and compare the strength of
phylogenetic signal across datasets, but analytical tools for these
comparisons are generally lacking. Here we develop a standardized effect
size, \(Z_\kappa\), which measures the strength of phylogenetic signal
on a common statistical scale. We also provide a test statistic,
\(\hat{Z}_{12}\), for comparing the strength of phylogenetic signal
across datasets. We find that two commonly used parameters (Pagel's
\(\lambda\) and Blomberg's \(\kappa\)), not converted to effect sizes,
are unsuitable for this purpose. Our effect-size procedure enables
biologists to quantitatively address hypotheses that compare the
strength of phylogenetic signal between various phenotypic traits, even
when those traits are found in different evolutionary lineages or have
different units or scales.}

@@

\begin{abstract}
Macroevolutionary studies frequently characterize the phylogenetic
signal in phenotypes, however, analytical tools for comparing the
strength of that signal across traits remain largely underdeveloped.
Here we evaluate the efficacy of of Pagel's \(\lambda\) to correctly
estimate the strength of phylogenetic signal in phenotypic traits across
a range of input values. We find that \(\lambda\) behaves as a Bernoulli
random variable, where estimates are increasingly skewed at larger and
smaller input levels of phylogenetic signal. Further, the precision of
\(\lambda\) varies with input signal. Another measure, Blomberg's
\(\kappa\), is more consistent across a range of tree sizes, and
exhibits a positive relationship with input levels of phylogenetic
signal. However, that relationship is decidedly nonlinear. Thus, neither
\(\lambda\) nor \(\kappa\) are suitable as effect sizes for measuring
the strength of phylogenetic signal, and comparing that signal across
datasets. As an alternative, we propose a standardized effect size based
on \(\kappa\), (\(Z_\kappa\)), which measures the strength of
phylogenetic signal more reliably than does \(\lambda\), and places that
signal on a common scale for statistical comparison. We develop tests
based on \(Z_\kappa\) to provide a mechanism for formally comparing the
strength of phylogenetic signal across datasets, in much the same manner
as effect sizes may be used to summarize patterns in quantitative
meta-analysis. Our approach extends the phylogenetic comparative toolkit
to address hypotheses that compare the strength of phylogenetic signal
between various phenotypic traits, even when those traits are found in
different evolutionary lineages or have different units or scales.
\end{abstract}

@@

Investigating macroevolutionary patterns of trait variation requires a
phylogenetic perspective, because the shared ancestry among species
violates the assumption of independence among trait values that is
common for statistical tests (1, 2). Accounting for this evolutionary
non-independence is the purview of \emph{phylogenetic comparative
methods} (PCMs): a suite of analytical tools that condition trends in
the data on the phylogenetic relatedness of observations (3--10). These
methods are predicated on the notion that phylogenetic signal -- the
tendancy for closely related species to display similar trait values --
is present in cross-species datasets (1, 11, 12). Indeed, under numerous
evolutionary models, phylogenetic signal is to be expected, as
stochastic character change along the hierarchical structure of the tree
of life generates trait covaration among related taxa (1, 12, 13).

Several analytical tools have been developed to quantify phylogenetic
signal in phenotypic datasets (11, 12, 14--17), and their statistical
properties -- namely type I error rates and statistical power -- have
been investigated to determine under what conditions phylogenetic signal
can be detected (13, 16, 18--23). One of the most widely used methods
for characterizing phylogenetic signal is Pagel's \(\lambda\) (11),
which transforms the lengths of the internal branches of the phylogeny
to improve the fit of data to the phylogeny via maximum likelihood (11,
24). When incorporated in PGLS, \(\lambda\) serves as a tuning parameter
which is optimized via log-likelihood profiling while evaluating the
covariation between the dependent and independent variables, given the
phylogeny (11, 24). To infer whether phylogenetic signal differs from no
signal or a Brownian motion model of evolutionary divergence, the
observed model fit using \(\hat\lambda\) may be statistically compared
to that using \(\lambda=0\) or \(\lambda=1\) via likelihood ratio tests
(24--26) or confidence limits (27).

Another widely used measure of phylogenetic signal is Blomberg's
\(\kappa\) (12), which characterizes phylogenetic signal as the ratio of
observed trait variation to the amount of variation expected under
Brownian motion. Blomberg's \(\kappa\) can be treated as a test
statistic by employing a permutation test to generate its sampling
distribution (12, 16) for determining whether significant phylogenetic
signal is present in data. Both \(\lambda\) and \(\kappa\) seem
intuitive to interpret, as a value of \(0\) for both corresponds to no
phylogenetic signal, while a value of \(1\) corresponds to the amount of
phylogenetic signal expected under Brownian motion. Thus, it is tempting
to regard both \(\lambda\) and \(\kappa\) as descriptive statistics that
measure the relative strength of phylogenetic signal, providing an
estimate of its magnitude for comparison.

The appeal of Pagel's \(\lambda\) and Blomberg's \(\kappa\) as
descriptive statistics is that they provide a basis for interpreting
``weak'' versus ``strong'' phylogenetic signal; i.e., small versus large
values of \(\hat{\lambda}\) or \(\kappa\), respectively, in a
comparative sense (28--30). Nonetheless, an important question that has
yet to be considered is whether these statistics are or can be converted
to effect sizes for comparative analyses across datasets? To be
statistics representing phylogenetic signal, they should have reliable
distributional properties, which could be revealed with simulation
experiments. For instance, as a proportional random variable bounded by
\(0\) and \(1\), we might expect that \(\hat{\lambda}\) follows a
Bernoulli distribution (\textbf{add ref}); i.e., branch lengths in a
tree are scaled proportionally to the probability that data arise from a
BM process. Given a known \(\lambda\) value used to generate random data
on a tree, we would also expect that the mean of an empirical sampling
distribution of \(\hat{\lambda}\) would approximately equal \(\lambda\);
the dispersion of \(\hat{\lambda}\) would be largest at intermediate
values of \(\lambda\), \(\hat{\lambda}\) would be predictable over the
range of \(\lambda\) with respect to tree size; the distribution of
\(\hat{\lambda}\) would be symmetric at intermediate values of
\(\lambda\) and more skewed toward values of 0 or 1; and that the
distribution of \(\hat{\lambda}\) will be more platykurtic at
intermediate values of \(\lambda\), becoming more leptokurtic toward 0
and 1 (\textbf{add same ref}). Prior work (18) seems to support some of
these conjectures, based superficially on statistical moments for a
given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of
ref. (18)). However, because the ``strength of Brownian motion'' was
simulated as a varied weighted-average of data simulated on trees with
\(\lambda=0\) and \(\lambda=1\) and not as prescribed values of
\(\lambda\) (18), interpretation of these patterns is challenging.

By contrast, for Blomberg's \(\kappa\), which is positively ubounded, we
might expect that for any \(\lambda\) used to generate data, estimates
of \(\kappa\) might follow a normal distribution, with values
distributed symmetrically about the input value. This attribute seemed
less reasonable based on the simulations performed by Münkemüller et
al.~(18), which suggested that distributions were positively skewed and
that Blomberg's \(\kappa\) might not behave as a statistic that follows
a normal distribution. However, because their simulations used a
weighted combination of simulated phylogenetic signal strengths, strong
inferences are not possible (and distributional attributes were not the
intended result of their simulations). Thus, for both Pagel's
\(\lambda\) or Blomberg's \(\kappa\), evaluation of statistical moments
across a range of \(\lambda\) used to generate data would be valuable
for adjudicating the relibalilty of these statistics. Furthermore, these
are statistics that appear to have expected values that vary with tree
size (18), making comparisons across studies challenging. Therefore,
transformation of these statistics into \(Z\)-scores would allow
evaluation of the efficacy of each statistic to yield effect sizes that
could be used for comparisons of the strength of phylogenetic signal
across traits and lineages.

Here we use simulation experiments to compare the distributional
attributes of \(\hat{\lambda}\) and \(\kappa\), plus their effect sizes
(\(Z\)-scores), across a range of tree size and phylogenetic signal
strength. We find that estimates of \(\hat{\lambda}\) are increasingly
skewed at larger and smaller input levels of phylogenetic signal and at
smaller tree sizes, vary widely for a given input value of \(\lambda\),
and that the precision of \(\hat{\lambda}\) is not constant across its
range. By contrast, estimates of \(\kappa\) are more consistent across
tree sizes, and are normally distributed across the range of input
levels of \(\lambda\), making \(\kappa\) a more reliable statistic. We
then propose an effect size based on \(\kappa\), (\(Z_{\kappa}\)), which
provides consistent estimates of the strength of phylogenetic signal
across tree sizes and signal strength, and facilitates quantitative
comparisons of the relative strength of phylogenetic signal across
datasets.

@@

\textbf{Lambda (\(\lambda\)) estimates of phylogenetic signal are
innacurate.} Computer simulations reveal that for \(\hat{\lambda}\), the
distributional expectations of a Bernoulli variable were mostly upheld.
First, the mean value of \(\hat{\lambda}\) increases as \(\lambda\)
increases, but it is negatively-biased (particularly for small tree
sizes), and is consistently less than the input \(\lambda\) value across
most of its range (Fig. 1 black line). Second, the standard deviation of
\(\hat{\lambda}\) is largest at intermediate values of \(\lambda\) and
smallest at extreme values, implying that the precision in estimating
\(\lambda\) varies across the range of input values (Fig. 1 red line).
Additionally, standard deviations of \(\hat{\lambda}\) are negatively
associated with tree size, and for tress of 128 species or less,
\(\hat{\lambda}\) are quite variable, except for cases when \(\lambda\)
is near or equal to \(1\). Third, the distributions of \(\hat{\lambda}\)
are not normal across its range, but become increasingly skewed at more
extreme values of \(\lambda\) (Fig. 1 blue line). For small tree sizes,
it is also clear that distributions are more platykurtic at intermediate
values of \(\hat{\lambda}\). Taken together these results reveal that
\(\hat\lambda\) inconsistently estimates phylogenetic signal, both
across tree sizes and across the range of input values. Additional
simulations (Supplemental Information) reveal that incorporating
\(\hat\lambda\) in PGLS anova and regression does not adversely affect
the statistical properties of PGLS parameter estimation or model
evaluation (type I error, power, bias in coefficients). Thus, it is
reasonable to incorporate \(\hat\lambda\) in PGLS as a parameter for
tuning the degree of phylogenetic signal in the dependent variables
during the analysis. However, the statistical properties shown in Fig. 1
demonstrate that \(\lambda\) is unsuitable as an effect size for
measuring the strength of phylogenetic signal in data, or for comparing
that signal across datasets.

\begin{figure}
\centering
\includegraphics{new.fig.1.temp.png}
\caption{Response of Pagel's \(\lambda\) to increasing strength of
Brownian motion. Gray line signifies the 1:1 line where the input value
matches the estimate \(\hat\lambda\). At each input level, the dark
black line represents the empirically derived expected value (mean) of
\(\hat\lambda\), the red line is the standard deviation of
\(\hat\lambda\), and the blue line is Shapiro Wilks statistic of
\(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent
skewed distributions).{}}
\end{figure}

\textbf{Kappa (\(\kappa\)) estimates of phylogenetic signal are more
reliable.} Simulation results for \(\hat\kappa\) demonstrate that
\(\hat\kappa\) displays better statistical properties. First, as
expected, mean values of \(\hat\kappa\) increase with increasing signal
(\(\lambda\)) irrespective of tree size, though the increase does not
scale linearly with input levels of phylogenetic signal (Fig. 2 black
line). Additionally, the standard deviation of \(\hat\kappa\) is
consistent across tree sizes (Fig. 2 red line), and while it increases
with \(\lambda\), it is always less than the corresponding mean. This
finding is perhaps unsurprising, as \(\hat\kappa\) is lower-bounded by
0, and is never large for small values of \(\lambda\). Importantly,
\(\hat\kappa\) is normally distributed across the range of input
\(\lambda\), and remains consistent in this pattern regardless of tree
size (Fig. 2 blue line). In fact, distributions of \(\hat\kappa\)
display slight skewing only at small tree sizes and when \(\lambda\) is
near or equal to \(1\). This result differs from those of (18), where
skewing appears to be from combining random values generated
independently. These findings reveal that while \(\kappa\) is more
reliable as an estimate of phylogenetic signal, the non-linear scaling
with input signal implies that it should not be considered an effect
size that measures the strength of phylogenetic signal on a common scale
for comparison across datasets.

\begin{figure}
\centering
\includegraphics{new.fig.2.temp.png}
\caption{Response of Blomberg's \(\hat\kappa\) to increasing strength of
Brownian motion. Gray line signifies the 1:1 line where the input value
matches the estimate \(\hat\kappa\). At each input level, the dark black
line represents the empirically derived expected value (mean) of
\(\hat\kappa\), the red line is the standard deviation of
\(\hat\kappa\), and the blue line is Shapiro Wilks statistic of
\(\hat\kappa\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent
skewed distributions).{}}
\end{figure}

\textbf{Effect sizes from \(\kappa\) (\(Z_{\kappa}\)) better
characterize phylogenetic signal.} To measure the strength of
phylogenetic signal on a common scale, here we propose effect sizes
(Z-scores) for both \(\lambda\) and \(\kappa\). Statistically, a
standardized effect size may be found as:

where \(\theta_{obs}\) is the observed test statistic, \(E(\theta)\) is
its expected value under the null hypothesis, and \(\sigma_\theta\) is
its standard error (31--33). Typically, \(\theta_{obs}\) and
\(\sigma_\theta\) are estimated from the data, while \(E(\theta)\) is
obtained from the distribution of \(\theta\) derived from parametric
theory. However, recent advances in resampling theory (34--37) have
shown that \(E(\theta)\) and \(\sigma_\theta\) may also be obtained from
an empirical sampling distribution of \(\theta\) obtained from
permutation procedures.

Formalizing the suggestion of Adams and Collyer (38), an effect size for
\(\kappa\) may be found as:

For \(\lambda\), deriving an effect size is more challenging, as
\(\lambda\) does not have a sampling distribution from which the
standard error and confidence intervals may be obtained. Confidence
intervals are therefore generated for the values of \(\lambda\) that
intersect the log-likehihood profile for corresponding percentiles of
the \(\chi^2\) distribution used to compare the putative model to a null
model with \(\lambda = 0\) {[}\textbf{add ref: MLC thinks Boettiger
paper?}{]}. Thus, an effect size for \(\lambda\) may be found as:

where, \(\hat{\lambda}\) is the maximized likelihood value of
\(\lambda\) and \(\chi^{2}_{\hat{\lambda}}\) is the likelihood ratio
statistic for the value.

Here we evaluate the ability of \(Z_{\lambda}\) and \(Z_{\kappa}\) to
characterize known levels of phylogenetic signal. Both \(Z_{\lambda}\)
and \(Z_{\kappa}\) are associated with input phylogenetic signal
(\(\lambda\)), indicating that both statistics capture the observed
signal (Fig. 3). However, effect sizes from \(\hat{\lambda}\) made
little sense, as they are more strongly associated with tree size than
they are with the actual phylogenetic signal in the data (Fig. 3). By
contrast, \(Z_{\kappa}\) is much more consistent accross tree sizes, and
increases more linearly with incrasing levels of phylogenetic signal.
Additionally, \(Z_{\kappa}\) exhibits a much stronger association with
phylogenetic signal strength as compared to tree size (Fig. 3), and its
standard deviation across input signal is more consistent. This implies
that similar levels of precision are found with \(Z_{\kappa}\) across
the range of input values. Thus between the two statistics,
\(Z_{\kappa}\) is a more reliable measure of the strength of
phylogenetic signal, and may be used to compare levels of phylogenetic
signal across datasets.

\begin{figure}
\centering
\includegraphics{new.fig.3.temp.png}
\caption{Response of effect sizes \(Z_\lambda\) and \(Z_\kappa\) to
increasing strength of Brownian motion.{}}
\end{figure}

\textbf{A test statistic (\(\hat{Z}_{12}\)) allows meaningful
comparisons across datasets.} To statistically compare the strength of
phylogenetic signal across datasets we propose a two-sample test
statistic (\(\hat{Z}_{12}\)). Based on statistical theory, a two-sample
test statistic may be calculated as:

where \(\kappa_1\), \(\kappa_2\), \(\hat\mu_{\kappa_1}\),
\(\hat\mu_{\kappa_2}\), \(\hat\sigma_{\kappa_1}\), and
\(\hat\sigma_{\kappa_2}\) are as defined above for equation 2. The right
side of the equation illustrates that if \(Z_\kappa\) has already been
calculated for two sampling distributions as in equation 2, the sampling
distributions have unit variance for each of the \(Z_\kappa\)
statistics. Estimates of significance of \(\hat{Z}_{12}\) may be
obtained from a standard normal distribution. Typically,
\(\hat{Z}_{12}\) is considered a two-tailed test, however directional
(one-tailed) tests may be specified should the empirical situation
require it (35, 37).

To demonstrate the utility of \(\hat{Z}_12\), we compared \(Z_{\kappa}\)
for two ecologically-relevant traits in plethodontid salamander (Fig.
4): surface area to volume ratios (SA:V) and relative body width
(\(\frac{BW}{SVL}\)) (39, 40). While both traits contained significant
phylogenetic signal, tests based on \(\hat{Z}_{12}\) revealed that the
degree of phylogenetic signal was significantly stronger in SA:V
(\(\hat{Z}_{12}=4.13\); \(P=0.000036\): Fig. 4). Biologically, this
observation may be interpreted by the fact that tropical species --
which form a monophyletic group within plethodontids -- display greater
variation in SA:V, which covaries with disparity in their climatic
niches (40). Thus, greater phylogenetic signal in SA:V is to be
expected.

@@

It is common in comparative evolutionary studies to characterize the
phylogenetic signal in phenotypic traits to determine the extent to
which shared evolutionary history has generated trait covariation among
taxa. However, while numerous analytical approaches may be used to
quantify phylogenetic signal (11, 12, 14--16), methods that explicitly
measure the strength of phylogenetic signal, or facilitate comparisons
among datasets, have remained underdeveloped. We evaluated the precision
of one common measure, Pagel's \(\lambda\), and explored its efficacy
for characterizing the strength of phylogenetic signal in phenotypic
data. Using computer simulations, we found that \(\lambda\) behaves as a
Bernoulli random variable, with estimates that are increasingly skewed
at larger and smaller input levels of phylogenetic signal. Further, the
precision of \(\lambda\) in estimating actual levels of phylogenetic
signal varies with both tree size (see also ref. (23)) and input levels
of phylogentetic signal. From these findings we conclude that
\(\lambda\) is not a reliable indicator of the observed strength of
phylogenetic signal in phenoytpic datasets, and should not be used as an
effect size for comparing the degree of phylogenetic signal between
datasets.

As an alternative, we described a standardized effect size (\(Z\)) for
assessing the strength of phylogenetic signal. \(Z\) expresses the
magnitude of phylogenetic signal as a standard normal deviate, which is
easily interpretable as the strength of phylogenetic signal relative to
the mean. We applied this concept to both \(\lambda\) and \(\kappa\),
and found that \(Z_\kappa\) was a better estimate of the strength of
phylogenetic signal in phenotypic data. First, values of \(Z_\kappa\)
more accurately tracked known changes in the magnitude of phylogenetic
signal, as demonstrated by the linear relationship between \(Z_\kappa\)
and input signal. Additionally, the precision of \(Z_{\kappa}\) was more
consistent across the range of input levels of phylogenetic signal.
Thus, \(Z_\kappa\) is a more reliable measure of the relative strength
of phylogenetic signal, and places that effect on a common and
comparable scale. We therefore recommend that future studies interested
in evaluating the strength of phylogenetic signal incorporate
\(Z_\kappa\) as a statistical measure of this effect.

Next we proposed a two-sample test (\(\hat{Z}_{12}\)), which provides a
formal statistical procedure for determining whether the strength of
phylogenetic signal is greater in one phenotypic trait as compared to
another. Prior studies have summarized patterns of variation in
phylogenetic signal across datasets using summary test values, such as
\(\kappa\) (12). However, because \(\kappa\) does not scale linearly
with input levels of phylogenetic signal (Fig. 2), and its variance
increases with increasing strength of phylogenetic signal (18, 20), it
should not be considered an effect size that measures the strength of
phylogenetic signal on a common scale. By contrast, standardizing
\(\kappa\) to \(Z_\kappa\) via equation 2 alleviates these concerns, and
facilitates formal statistical comparisons of the strength of signal
across datasets. Thus when viewed from this perspective, the approach
developed here aligns well with other statistical approaches such as
meta-analysis (31, 41, 42), where summary statistics across datasets are
converted to standardized effect sizes for subsequent ``higher order''
statistical summaries or comparisons. As such, our approach enables
evolutionary biologists to quantitatively examine the relative strength
of phylogenetic signal across a wide range of phenotypic traits, and
thus opens the door for future discoveries that inform on how phenotypic
diversity accumulates in macroevolutionary time across the tree of life.
